Parallel Adaptive Importance Sampling

### 2016-06-19 PAIS.tex
corrections in the rest of the sections

In this example we compare the naively parallelised RWMH algorithm with its
PAIS variant, the PAIS-RW algorithm. We perform this comparison using the relative $L^2$ error defined in Equation~\ref{eqn:L2_error}, as well as the relative error in the first moment.

Since we are comparing against naively parallelised MH algorithms,
we also need to decide which statistic $T(\beta)$ to optimise for
these approaches. In the examples which follow, we have optimised the
naively parallelised RWMH algorithm using the optimal acceptance rate
$\hat{\alpha} = 0.5$. This value differs from the theoretical
asymptotic value of 0.234 which applies in higher dimensions, but this higher acceptance rate is commonly used for one dimensional Gaussian posteriors~\cite{rosenthal2011optimal}. To optimise the algorithm we minimise the statistic
\[
	T_{\text{MH}}(\beta) = \left| \frac{N_{\text{acc}}(\beta)}{N_{\text{total}}} - \hat{\alpha} \right|,
\]
where $N_{\text{acc}}(\beta)$ is the number of accepted moves and
$N_{\text{total}}$ is the total number of samples produced. For the
PAIS algorithm, we are maximising the effective sample size as
discussed in Section~\ref{sec:ess}.

@@

Consider the simple case of a linear observation operator $\G(u) = u$, where the prior on $u$ and the observational noise follow Gaussian distributions. Then, following Equation~\ref{eqn:like}, the Gaussian posterior has the form
\begin{equation}\label{eqn:Gaussian posterior}
	\text{law}(\mu_D) = \pi(u|D) \propto \exp\left(-\frac{1}{2}\big\|u - D\big\|^2_{\sigma^2} - \frac{1}{2}\big\|u\big\|^2_{\tau^2}\right),
\end{equation}
where $\sigma^2$ and $\tau^2$ are the variances of the
observational noise and prior distributions respectively. In the
numerics which follow, we choose $\tau^2 =0.01$ and $\sigma^2 = 0.01$, and
we observe $u=4$ noisily such that
\[
D = \mathcal{G}(u) + \eta \sim \mathcal{N}(\mathcal{G}(u),\sigma^2) = \mathcal{N}(4,0.01).
\]
These values result in a posterior density in which the vast majority
of the density is out in the tails of the
prior distribution. The Kullback-Leibler (KL)
divergence, which gives us a measure of how different the prior and
posterior are, is $D_{KL}(\mu_D || \mu_0) = 4.67$ for this
problem. A KL divergence of zero indicates that two distributions are
identical almost everywhere.

@@

{\bf (3) Measuring convergence of adaptive algorithms}: We run the
adaptive algorithms under the same conditions as the nonadaptive
algorithms, and again use the relative $L^2$ error to compare
efficiency. The adaptive algorithms are initialised with $\beta^{(1)}$ = 2.

@@

Both adaptive algorithms are run with initial values of $\beta=2.0$. Figure~\ref{fig:MH1 L2} shows that after an initial burn-in period the APAIS-RW algorithm catches up to the PAIS-RW algorithm, and by the end of the simulation window is matching its performance. The ARWMH algorithm does not perform quite as well, this is possibly due to the fact that the acceptance rate cost function is not smooth at the optimal value making it difficult to minimise.

@@

Figure~\ref{fig:PAIS_saving} was produced using the method of finding
optimal $\beta$ described in Section~\ref{sec:Implementation P1} (1),
then running 32 repeats at each ensemble size. The convergence rates
are then found by regressing through the data. The graph is still very
noisy but demonstrates that increasing the ensemble size continues to
reduce the number of iterations required in comparison with naively parallelised MH. When $M<8$, the algorithm takes a long time to reach
stationarity. The decreasing trend indicates superlinear improvement of PAIS with
respect to ensemble size, in terms of the number of iterations
required, which is a demonstration of our belief that parallelism of
MCMC should give us added value over and above that provided by naive parallelism. This decrease is linked to the increasing effective sample size shown in Figure~\ref{fig:neff-M} (b).

@@

The following setup is the same for both problems. We consider a nonlinear observation operator $\G(u) = u^2$, and assign the prior $u \sim \mu_0 = \N(0, \tau^2=0.25)$. We assume that a noisy reading, $D$, is taken according to $D = \G(u) + \varepsilon$, where $\varepsilon \sim \mu_\varepsilon = \N(0, \sigma^2 = 0.1)$. This results in the non-Gaussian posterior
\[
	\pi(u|D) \propto \exp\left(-\frac{1}{2\sigma^2}\|u^2 - D\|^2 - \frac{1}{2\tau^2}\|u\|^2\right).
\]
To create the `easy' problem we say that the true value of $\G(u) = 0.75$, and the `hard' problem is generated using $\G(u) = 2$. In the numerics which follow we draw noise from $\mu_\varepsilon$ to generate our data point.

@@

Calculating the optimal values of the scaling parameters for this
problem is similar to the previous example; we check only the acceptance rate to find the optimal values for RWMH and we use the effective sample size to find the optimal values for PAIS-RW. Table~\ref{table:BM_opt_beta} gives the optimal values of $\beta$ for both problems.

@@

It is relatively simple to find optimal scaling parameters for problem $B_1$. These values are given in Table~\ref{table:BM_opt_beta} (left). However problem $B_2$ is much harderas transitions between the modes are extremely unlikely for the standard RWMH algorithm. This means that we need to consider the
convergence on two levels; we should consider the algorithm's ability
to find all the modes, and also whether it can sample them in the correct proportions.

To get correctly proportioned modes with the RWMH algorithm it is
important that the chains can transition between the modes, which
means that $\beta$ must be large. However, this leads to a lower
acceptance rate, and so we sacrifice convergence locally. For this reason, the RWMH algorithm is very slow to
converge for problems of this type.

@@

If we assume that we know the rate constants $k_1$ and $k_4$, and
observe the system in Equation~\ref{eqn:QSSA_system}, we indirectly
observe the rates $k_2$ and $k_3$ through the effective rate
$\hat{k}_4$ of the degradation of $\mathcal{S}$. Our
observations are uninformative about these reaction rates, as there
are surfaces in parameter space along which the effective rate
$\hat{k}_4$ is invariant, leading to a highly illposed inverse
problem. Making the assumption that the errors in our observations of
the value of $\mathcal{S}$ are Gamma distributed with some variance
$\sigma^2$, this results in a long and thin posterior distribution on
$k_2$ and $k_3$. This type of problem is notoriously difficult to
sample from using standard MH algorithms, as the algorithms quickly
find a point on this manifold on which $\hat{k}_4$ is invariant, but
exploration along its length is slow.

@@

\subsubsection{Implementation}\label{sec:chem_implementation}
% explanation of l2 convergence measure
For this problem there is no analytic form for the normalisation
constant, and numerical methods implemented in MATLAB and Mathematica
have proven to be unreliable. To demonstrate the convergence of the
algorithm, we first run a MH algorithm for much longer than we
normally would ($8\times 10^{10}$ samples), and consider the histogram
produced to be a well-converged approximation of the true posterior
distribution. We then perform our usual simulations with the PAIS and
MH algorithms to compare the rate at which the histograms converge to
the posterior over the same histogram bins.

>>>>

In this example we compare the naively parallelised RWMH algorithm with its
PAIS variant, the PAIS-RW algorithm. We asses the performance of the PAIS algorithm
using the relative $L^2$ error defined in Equation~\ref{eqn:L2_error}, as well as the
relative error in the first moment.

Since we are comparing against naively parallelised MH algorithms,
we also need to decide which statistics $T(\beta)$ provide the best criterions
for obtaining the optimal scaling parameters. In the examples which follow, we have
optimised the naively parallelised RWMH algorithm using the optimal
acceptance rate $\hat{\alpha} = 0.5$. This value differs from the theoretical
asymptotic value of 0.234 which applies in higher dimensions, but this higher acceptance rate is commonly used for one dimensional Gaussian posteriors~\cite{rosenthal2011optimal}. To find the optimal
scaling parameter we minimise the statistic
\[
	T_{\text{MH}}(\beta) = \left| \frac{N_{\text{acc}}(\beta)}{N_{\text{total}}} - \hat{\alpha} \right|,
\]
where $N_{\text{acc}}(\beta)$ is the number of accepted moves and
$N_{\text{total}}$ is the total number of samples produced. For the
PAIS algorithm, we maximise the effective sample size as
discussed in Section~\ref{sec:ess}.

@@

Consider the simple case of a linear observation operator $\G(x) = x$, where the prior on $x$ and the observational noise follow Gaussian distributions. Then, following Equation~\ref{eqn:like}, the Gaussian posterior has the form
\begin{equation}\label{eqn:Gaussian posterior}
	\text{law}(\mu_D) = \pi(x|D) \propto \exp\left(-\frac{1}{2}\big\|x - D\big\|^2_{\sigma^2} - \frac{1}{2}\big\|x\big\|^2_{\tau^2}\right),
\end{equation}
where $\sigma^2$ and $\tau^2$ are the variances of the
observational noise and prior distributions respectively. In the
numerics which follow, we choose $\tau^2 =0.01$ and $\sigma^2 = 0.01$, and
we observe $x_\text{ref}=4$ noisily such that
\[
D = \mathcal{G}(x) + \eta \sim \mathcal{N}(\mathcal{G}(x_\text{ref}),\sigma^2) = \mathcal{N}(4,0.01).
\]
These values result in a posterior density in which the vast majority
of the density is out in the tails of the
prior distribution. The Kullback-Leibler (KL)
divergence, which gives us a measure of how different the prior and
posterior are, is $D_{KL}(\mu_D || \mu_0) = 4.67$ for this
problem. A KL divergence of zero indicates that two distributions are
identical almost everywhere.

@@

{\bf (3) Measuring convergence of adaptive algorithms}: We run the
adaptive algorithms under the same conditions as the nonadaptive
algorithms, and again use the relative $L^2$ error to compare
efficiency. The adaptive algorithms are initialised with $\beta^{(1)}$ = 1.

@@

Both adaptive algorithms are run with initial values of $\beta=1$. Figure~\ref{fig:MH1 L2} shows that after an initial burn-in period the APAIS-RW algorithm catches up to the PAIS-RW algorithm, and by the end of the simulation window is matching its performance. The ARWMH algorithm does not perform quite as well, this is possibly due to the fact that the acceptance rate cost function is not smooth at the optimal value making it difficult to minimise.

@@

Figure~\ref{fig:PAIS_saving} was produced using the method of finding
optimal $\beta$ described in Section~\ref{sec:Implementation P1} (1),
then running 32 repeats at each ensemble size. The convergence rates
are then found by regressing through the data. The graph is still very
noisy but demonstrates that increasing the ensemble size continues to
reduce the number of iterations required in comparison with naively
parallelised MH. The decreasing trend indicates superlinear improvement of PAIS with
respect to ensemble size, in terms of the number of iterations
required, which is a demonstration of our belief that parallelism of
MCMC should give us added value over and above that provided by naive parallelism. This decrease is linked to the increasing effective sample size shown in Figure~\ref{fig:neff-M} (b).

@@

The following setup is the same for both problems. We consider a
non-linear observation operator $\G(x) = x^2$, and assign the prior
$x \sim \mu_0 = \N(0, \tau^2=0.25)$. We assume that a noisy reading,
$D$, is taken according to $D = \G(x_\text{ref}) + \varepsilon$, where
$\varepsilon \sim \mu_\varepsilon = \N(0, \sigma^2 = 0.1)$. This results
in the non-Gaussian posterior
\[
	\pi(x|D) \propto \exp\left(-\frac{1}{2\sigma^2}\|x^2 - D\|^2 - \frac{1}{2\tau^2}\|x\|^2\right).
\]
To create the `easy' problem we say that the true value of
$\G(x_\text{ref}) = 0.75$, and the `hard' problem is generated using
$\G(x_\text{ref}) = 2$. In the numerics which follow we draw noise from
$\mu_\varepsilon$ to generate our data point.

@@

Calculating the optimal values of the scaling parameters for this
problem is similar to the previous example; we check only the acceptance
rate to find the optimal values for RWMH and we use the effective
sample size to find the optimal values for PAIS-RW.
Table~\ref{table:BM_opt_beta} gives the optimal values of $\beta$ for both
problems. The subscript on $\beta$ refers to the criterion which has been optimised.

@@

It is relatively simple to find optimal scaling parameters for problem $B_1$.
These values are given in Table~\ref{table:BM_opt_beta} (left). However
problem $B_2$ is much harder as transitions between the modes are extremely
unlikely for the standard RWMH algorithm. This means that we need to consider the
convergence on two levels; we should consider the algorithm's ability
to find both the modes, and also whether it can sample them in the correct proportions.

To get correctly proportioned modes with the RWMH algorithm it is
important that the chains can transition between the modes frequently, which
means that $\beta$ must be large. However, this leads to a lower
acceptance rate, and so we sacrifice convergence locally. For this reason, the RWMH algorithm is very slow to
converge for problems of this type.

@@

If we assume that we know the rate constants $k_1$ and $k_4$, and
observe the system in Equation~\ref{eqn:QSSA_system}, we indirectly
observe the rates $k_2$ and $k_3$ through the effective rate
$\hat{k}_4$ of the degradation of $\mathcal{S}$. Our
observations are uninformative about these reaction rates, as there
are surfaces in parameter space along which the effective rate
$\hat{k}_4$ is invariant, leading to a highly ill-posed inverse
problem. Making the assumption that the errors in our observations of
the value of $\mathcal{S}$ are Gamma distributed with some variance
$\sigma^2$, this results in a long and thin posterior distribution on
$k_2$ and $k_3$. This type of problem is notoriously difficult to
sample from using standard MH algorithms, as the algorithms quickly
find a point on this manifold on which $\hat{k}_4$ is invariant, but
exploration along its length is slow.

@@

\subsubsection{Implementation}\label{sec:chem_implementation}
% explanation of l2 convergence measure
For this problem there is no analytic form for the normalisation
constant, and numerical methods implemented in MATLAB and Mathematica
have proven to be unreliable. To demonstrate the convergence of the
algorithm, we first run a MH algorithm for much longer than we
normally would ($8\times 10^{10}$ samples), and consider the histogram
produced to be a well-converged approximation of the true posterior
distribution. We then perform our usual simulations with the PAIS and
MH algorithms to compare the rate at which the histograms converge to
the posterior over the same mesh.
